Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame

### 2023-03-14 manuscript/Chapter2-manuscript.Rmd
first draft of discussion complete

Hunting experience is predicted to imbue an increase in predator foraging specialization when resources are predictable [@Estes.etal2003;@Tinker.etal2008;@Tinker.etal2009]. Other lines of research view learning from experience as a form of behavioural plasticity [@Ishii.Shimada2010;@Mery.Burns2010;@Snell-Rood2013;@Wright.etal2022], with theoretical models predicting an increase in flexibility from experience when resources are variable [@Stephens1993]. Our results challenge the hypothesis that experience should generate a directional population-level increase in behavioural specialization or plasticity. Instead, the consequences from experience appear to be a property of individuals, such that each predator developed, through extensive practice, their own expertise and behavioural trajectory; some specialized at fast-paced hunting, while others adopted a slower and more flexible tactic. The predators' tactics were matched to prey encounters, such that specialized cursorial hunters experienced more predictable encounters, while slower and flexible hunters experienced more unpredictable encounters, with no difference in success between the two predator-types. This predator-prey behavioural matching grew stronger with predator experience, while the relationship between the prey's behaviour and hunting success diminished. Together, our observations suggest that predators learn to adjust their hunting tactic with experience, and the degree at which they specialize in it, based on the behaviour of the prey that they encountered throughout their lifetime. Moreover, the behavioural escalation through experience suggests that learning may be a catalyst for predator-prey arms races.

@@

Empirical studies in humans and animals show that proficiency in a given task often increases nonlinearly with experience, and stabilizes when an expert level is reached [reviewed in @Dukas2019]. However, these patterns are derived from models that assume the same increase in expertise for each individual, and their data do not include repeated measures of individual success through time. In contrast with these studies, our analyses revealed that there was no population increase in expertise; instead, individual predators in *DBD* displayed distinct patterns of expertise acquisition. For instance, many predators increased their success (Figure S1A), but others displayed a decrease in success (Figure S1B) or no change in success at all with experience. We predicted that this would occur because predators should differ in the type of prey that they encounter, as some may be easier to capture than others. This is often the case in nature as prey can use variety of defenses such as physical armaments, toxins, camouflage, or antipredator behaviours to escape predation [@BrodieIII.BrodieJr.1999;@Bowen.etal2002;@brodieCoEvolutionPredatorsPrey2010;@Nomura.etal2011;@Carey.Wahl2011]. However, our results show that the speed of the prey did not shape individual differences in skill acquisition by predators, even if they differed in the prey that they encountered. This was unexpected, because prey speed is an important predictor of predator success in this system [@FraserFranco.etal2022] as well as in others [@Walker.etal2005;@Kelley.Magurran2011]. Despite that other antipredator tactics may mediate these individual responses by predators, it is likely that individual differences in motivation and perseverance, attention span, or memory retention were also major causes [@Warburton2003;@morand-ferronStudyingEvolutionaryEcology2016;@Dukas2019]. For instance, a persevering predator may increase its success much more steadily than an unmotivated predator that quickly abandons a chase. Another factor that may explain differences in the development of expertise is the time delay between hunting events, such that a predator that played 300 matches in the span of six months may not retain information (e.g. prey escape patterns or muscle memory) as well as one that played 300 matches in the span of 6 days. While this has not been formally tested, it is hypothesized that a greater delay between exposure events should attenuate or even negate the relationship between cumulative experience and success as a result of forgetting [@Endler1991;@Wright.etal2022].

@@

We detected a similar pattern for foraging specialization, such that individual differences increased with experience, while specialization at the population level remained stable. *parler du patron non linéaire et donner quelques potentielles explications liées à l'exploration + grande chez les juvéniles*. *Je peux citer les sources de Phillips et al : Thiers 2014 et Grissac 2016*.

Predators that experienced predictable encounters with their prey specialized in hunting at high speeds, while those that experienced more unpredictable encounters adopted a slower and flexible strategy. This is similar to other studies finding that fast-paced hunting is a highly specialized tactic suited for prey that use rapid escapes [@Endler1991;@Bro-Jorgensen2013;@Wilson.etal2018;@Szopa-Comley.Ioannou2022]. On the contrary, instead of focusing solely on one tactic, predators that experienced unpredictable encounters probably adjusted their behaviour to minimize the consequences of uncertainty. Interestingly, both specialist and flexible hunters fine-tuned their tactic to their prey as they gained experience, suggesting that they learned how to hunt their prey. In that regard, our results provide direct evidence that the mechanism underlying the resource-predictability hypothesis may indeed involve predators learning to adjust their degree of specialization based on their prey [@Weimerskirch.etal2005;@Weimerskirch2007;@Woo.etal2008;@Phillips.etal2017]. This is further supported by the fact that specialist and flexible hunters both achieved similar success throughout the study period. Therefore, our results have important implications for future predator-prey models, which should incorporate individual variation in learning and prey encounters to better predict under which ecological contexts specialization should be favoured over flexibility. *important de faire ressortir que c'est seulement 10% des gens qui ont vraiment changé leur comportement de beaucoup. Ainsi ça va être + nuancé et me permettre de dire que certains ne répondent pas à l'expérience*.*affirmer que pt qu'ils répondent à d'autres expériences, mais que des contraintes logistiques nous empêchent de savoir quoi*

```{r eval = FALSE}
*je peux commencer à plug le arms race dans ce paragraphe.*
*il faut aussi que je parle du fait que la variation dans le succès de capture a diminué avec l'XP*
*- Experience strengthens predator-prey interactions*
*- Experience dampens the relationship between behaviour and success*

Contrary to Hanifin et al. 2008, we find a directional symmetry (i.e. trait matching).
Our results are similar to Szopa-Comley and Ioannou 2022. Adjustments based on prey. Le predateur s'ajuste dans le traitement unpredictable.
In abrams 2000, we read : "The interaction of predator and prey traits in determining capture rate seems to be the most important factor determining the stability of the entire system. Most studies in which cycles are a commonly observed outcome assume that the predator maximizes its rate of capture of the prey by matching the prey's phenotype"
We can also read : "If there is sufficiently strong stabilizing selection based on the costs of extreme trait values in both species, then stable equilibria are likely"
In Reimche et al. 2020 : in the discussion we can read that we still don't understand the factors driving phenotypic (mis)matching. In my case, it appears that learning may play a role.
Revoir mes résultats dans le cadre des résultats de Glaudas et al. 2019. Les 2 concordent bien.
Voir Miszei comme exemple parmi d'autres pour trait matching.
check Mougi etal. 2011 for arms-races based on phenotypic plasticity!

@@

*Our results also have implications also for predator-prey models of trait-matching, because such models are usually studied at the evolutionary scale.*

>>>>

Hunting experience is predicted to imbue an increase in predator foraging specialization when resources are predictable [@Estes.etal2003;@Tinker.etal2008;@Tinker.etal2009]. Other lines of research view learning from experience as a form of behavioural plasticity [@Ishii.Shimada2010;@Mery.Burns2010;@Snell-Rood2013;@Wright.etal2022], with theoretical models predicting an increase in flexibility from experience when resources are variable [@Stephens1993]. Our results challenge the hypothesis that experience should generate a directional increase in behavioural specialization or plasticity in a population. Instead, the consequences from experience appear to be a property of individuals, such that each predator developed, through extensive practice, their own expertise and behavioural trajectory; some specialized at fast-paced hunting, while others adopted a slower and more flexible tactic. The predators' tactics were matched to prey encounters, such that specialized cursorial hunters experienced more predictable encounters, while slower and flexible hunters experienced more unpredictable encounters, with no difference in success between the two predator-types. This predator-prey behavioural matching grew stronger with predator experience, while the relationship between the prey's behaviour and hunting success diminished. Together, our observations suggest that predators learn to adjust their hunting tactic with experience, and the degree at which they specialize in it, based on the behaviour of the prey that they encountered throughout their lifetime. Moreover, the behavioural escalation through experience suggests that learning may be a catalyst for predator-prey arms races.

@@

Empirical research in humans and animals show that the proficiency in a given task often increases nonlinearly with experience and stabilizes at an expert level [reviewed in @Dukas2019]. However, these results are derived from models that assume the same increase in expertise for each individual, and the data from these studies does not include repeated measures of individual success through time. In contrast, our analyses revealed that there was no population increase in expertise; instead, individual predators in *DBD* displayed distinct patterns of expertise acquisition. For instance, many predators increased their success (Figure S1A), but others displayed a decrease in success (Figure S1B) or no change in success at all with experience. We predicted that this would occur because predators should differ in the type of prey that they encounter, as some may be easier to capture than others. This is often the case in nature as prey can use variety of defenses such as physical armaments, toxins, camouflage, or antipredator behaviours to escape predation [@BrodieIII.BrodieJr.1999;@Bowen.etal2002;@brodieCoEvolutionPredatorsPrey2010;@Nomura.etal2011;@Carey.Wahl2011]. However, the speed of the prey did not shape individual differences in expertise acquisition by predators, even if they differed in the prey that they encountered. This was unexpected, because prey speed is an important predictor of predator success in this system as well as in others [@Walker.etal2005;@Kelley.Magurran2011;@FraserFranco.etal2022]. Despite that other antipredator tactics may mediate these individual responses by predators, it is likely that individual differences in motivation and perseverance, attention span, or memory retention were also major causes [@Warburton2003;@morand-ferronStudyingEvolutionaryEcology2016;@Dukas2019]. For instance, a persevering predator may increase its success much more steadily than an unmotivated predator that quickly abandons a chase. Another factor that may explain differences in the development of expertise is the time delay between hunting events, such that a predator that played 300 matches in the span of six months may not retain information (e.g. prey escape patterns or muscle memory) as well as one that played 300 matches in the span of 6 days. While this has not been formally tested, it is hypothesized that a greater delay between exposure events should attenuate or even negate the relationship between cumulative experience and success, as a result of forgetting [@Endler1991;@Wright.etal2022].

@@

The predator population displayed an increase in foraging flexibility when they reached an intermediate level of experience. As they reached an advanced level of experience, they returned to a level of flexibility that was similar to when they were novices. This is comonly observed in juvenile predators across the animal kingdom, because exploring and learning different tactics at this developmental stage is crucial to become a skillful hunter [@Vehanen2003;@Johnson.Wilbrecht2011;@Thiers.etal2014;@deGrissac.etal2016]. Thus, the predators in *DBD* were probably exploring and refining different tactics at this stage of their development in the game. Interestingly, individuals became increasingly distinct in their level of foraging specialization with experience, which was related to the degree of variability in prey encounters. Predators that experienced predictable encounters with their prey specialized in hunting at high speeds, while those that experienced more unpredictable encounters adopted a slower and flexible strategy. This is similar to other studies finding that fast-paced hunting is a highly specialized tactic suited for prey that use rapid escapes [@Endler1991;@Bro-Jorgensen2013;@Wilson.etal2018;@Szopa-Comley.Ioannou2022]. On the contrary, instead of focusing solely on one tactic, predators that experienced unpredictable encounters probably adjusted their behaviour to minimize the consequences of uncertainty. As they gained experience, predators increasingly matched their tactic to their prey, suggesting that they learned how to hunt them throughout repeated interactions. In that regard, our results provide direct evidence that the mechanism underlying the resource-predictability hypothesis may indeed involve predators learning to adjust their degree of specialization based on their prey [@Weimerskirch.etal2005;@Weimerskirch2007;@Woo.etal2008;@Phillips.etal2017]. This is further supported by the fact that specialist and flexible hunters both achieved similar success throughout the study period.

An emerging pattern reported in predator-prey research is that predators and prey often match their phenotype to one another (i.e. positive trait covariances), which appears to be driven by an arms-race-like reciprocal phenotypic plasticity [@Kishida.etal2006;@Kishida.etal2009;@Edgell.Rochette2009a;@Mougi.etal2011;@McGhee.etal2013]. From an evolutionary perspective, @BrodieIII.BrodieJr.1999 showed in the garter snake (*Thamnophis sirtalis*) - roughskin newt (*Taricha granulosa*) system that such escalations may occur when prey develop defensive traits, which leads to stronger selection in predators, resulting in counteradaptations that circumvent the prey's defenses. In our system, the predator's success is negatively correlated to the prey's speed, which is an effective antipredator tactic [@FraserFranco.etal2022]. However, as the covariance between the predator's speed and the speed of the prey encountered increased with experience, reminiscent of an arms race, we found that the fitness consequences of the preys' speed on predators decreased with experience. Moreover, this pattern was followed by a decrease in among individual variation in success with experience. This has important implications for the role of experience on predator-prey interactions, as we found that learning enables predators to effectively develop countermeasures to the prey's defenses at nonevolutionary timescales. It also provides empirical evidence to the idea that reciprocal phenotypic plasticy through learning may attenuate selection exerted by prey on predators [@Anderson1995;@Ancel1999;@Borenstein.etal2006;@Paenke.etal2007].

## Conclusions

```{r eval = FALSE}

*duckworth en tete*

Contrary to Hanifin et al. 2008, we find a directional symmetry (i.e. trait matching).

In abrams 2000, we read : "The interaction of predator and prey traits in determining capture rate seems to be the most important factor determining the stability of the entire system. Most studies in which cycles are a commonly observed outcome assume that the predator maximizes its rate of capture of the prey by matching the prey's phenotype"
We can also read : "If there is sufficiently strong stabilizing selection based on the costs of extreme trait values in both species, then stable equilibria are likely"

In Reimche et al. 2020 : in the discussion we can read that we still dont understand the factors driving phenotypic (mis)matching. In my case, it appears that learning may play a role.
Revoir mes résultats dans le cadre des résultats de Glaudas et al. 2019. Les 2 concordent bien.

Voir Miszei comme exemple parmi dautres pour trait matching.

check Mougi etal. 2011 for arms-races based on phenotypic plasticity (modèle théorique)

voir Edgell : prey-induced plasticity leads to covariance between predator-prey trait

@@

Phrases de conclusion :

Our results also have implications also for predator-prey models of trait-matching, because such models are usually studied at the evolutionary scale.

Therefore, our results have important implications for future predator-prey models, which should incorporate individual variation in learning and prey encounters to better predict under which ecological contexts specialization should be favoured over flexibility. ---> pour la conclusion

important de faire ressortir que c'est seulement 10% des gens qui ont vraiment changé leur comportement de beaucoup. Ainsi ça va être + nuancé et me permettre de dire que certains ne répondent pas à l'expérience*.*affirmer que pt qu'ils répondent à d'autres expériences, mais que des contraintes logistiques nous empêchent de savoir quoi
